---
title: "Explore Neuro/Psychiatric PheWAS Codes"
author: 'Meg Hutch'
date: '2022-05-03'
output:
  html_document:
    toc: true
    toc_float: true
    code_download: true
    theme: spacelab
---

### **Evaluation of Neurological and Psychiatric PheWAS Codes and their association with PASC**

*Note: Ensure input_dir is changed to read your Phase2.2 LocalPatientObservations.csv file*

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(plotly)
library(DT)
library(stringr)
library(lubridate)
library(data.table)

siteid = "NWU"
input_dir = "/share/fsmresfiles/nm_cvd/4CE/PhaseX.2/Input/LocalPatientObservations.csv"
```

# Import Phase2.2 Data

We will import the Phase2.2 observations file and return only ICD-9/ICD-10 codes

```{r}
# ensure this directs to the input directory where Phase2.2 data files are stored
observations <- read.csv(input_dir)

observations <- observations %>% 
  filter(concept_type == 'DIAG-ICD10' | concept_type == 'DIAG-ICD9')
```

### Select COVID-19 positive patients

We will select only patients who had a positive PCR test

```{r}
positive <- c("Pos")

covid_pos <- observations %>% 
  filter(str_detect(cohort, paste(positive, collapse = "|")))
```

**Collapse patient cohorts regardless of time period**

We will create a `status` variable to identify patients who were"Admitted" vs "Not Admitted"

```{r}
covid_pos <- covid_pos %>% 
  mutate(status = if_else(str_detect(cohort, "NotAdm"), "Not Admitted", "Admitted"),
         covid = "Positive") %>% 
  select(patient_num, status, cohort, days_since_admission, concept_code, covid)

# total number of positive patients with codes>= 90 days after admission
pos_n = nrow(unique(data.frame(covid_pos$patient_num))); pos_n
table(covid_pos$status)

rm(observations)
```

# Create training & test sets

We will make sure our training & test sets are stratified equally by admission status

*Note: Make sure to run the whole block of code (set.seed) and groups together in order to maintain reproducibility*

**Randomly assign patients to train or test sets**

```{r}
set.seed(317)

groups <-
  covid_pos %>%
  distinct(patient_num, status) %>%
  rowwise() %>%
  mutate(split = sample(
    c("train", "test"),
    1,
    replace = FALSE,
    prob = c(0.8, 0.2) # Set weights for each group here
  )) %>% 
  data.frame()


# make sure dataset is balanced
nrow(covid_pos %>% distinct(patient_num))
nrow(covid_pos %>% filter(status == 'Admitted') %>% distinct(patient_num))
nrow(covid_pos %>% filter(status == 'Not Admitted') %>% distinct(patient_num))

# 6/14/2022: 6% pts are admitted
table(groups$status, groups$split)

# the math looks good to me here
```

**Create the train and test data frames**

```{r}
train <- covid_pos %>% 
  inner_join(., groups %>% 
               filter(split == "train") %>% 
               distinct(patient_num))
nrow(unique(data.frame(train$patient_num)))

test <- covid_pos %>% 
  inner_join(., groups %>% 
               filter(split == "test") %>% 
               distinct(patient_num))

nrow(unique(data.frame(test$patient_num)))
```

# Identify Neuro Patients in the Training Set

## Evaluate Harrison Zhang's codes

Harrison provided a list of PheCodes or "phenotypes whose new onset is conditionally dependent on exposure to the virus 90 or 180 days prior to new onset. Thus, we're identifying phenotypes that we believe are forms of PASC - thus, these codes could be considered a proxy for PASC".

We have subset this list to include phenotypes under the neurological and mental disorder groups. This list also contains a few "rescued" codes from other phenotype groups including sense organs, symptoms, and circulatory system, that Zongqi and Malar identified as also being neurologically related.

```{r}
# load in phecodes from HZ's analysis
neuro_phecodes <- read.csv("data/2022-04-13_dCRT_meta_logistic_beta_pos_results_unique_sig_codes_MS_ZX.csv") %>% 
  filter(`Include.Neuro.Pasc..1...yes..0...no.` == 1) %>% 
  select(phecode, description, group)

datatable(neuro_phecodes %>% arrange(phecode),
          filter = "top")
```

### Import All PheCodes

We will also add ICD-10 diagnosis codes to PheCode table to facilitate downstream merging with 4CE data

```{r}
# import the dataframe of all phecodes processed in R/process_all_phecodes.R
phecodes_all <- read.csv("data/all_phecodes.csv")

neuro_phecodes_icd <- neuro_phecodes %>% 
  left_join(., phecodes_all %>% 
              rename(Phenotype = "phecode_str",
                     `ICD10.String` = "icd10cm_str",
                     ICD10 = "icd10cm") %>% 
              select(phecode, Phenotype, ICD10, ICD10.String), 
            by = "phecode")
```

### Evaluate COVID+ patients with NeuroPASC codes within 90 days post-admission

This will select any patient who has a code at least 90 days after their initial hospitalization

```{r}
pasc_90_train_hz <- train %>%
  inner_join(., neuro_phecodes_icd %>%  
            rename(concept_code = "ICD10"), 
            by = "concept_code") %>% 
  filter(days_since_admission >= 90)
```

#### Frequency of PheCodes after 90 days

Evaluate the prevalence of each PheCode

```{r fig.height=16, fig.width=10}
pasc_90_hz <- pasc_90_train_hz %>% 
  group_by(Phenotype, status) %>% 
  mutate(count = n_distinct(patient_num),
         percent = round(count/nrow(train %>% distinct(patient_num))*100,2)) %>% 
  ungroup() %>% 
  distinct(phecode, Phenotype, count, group, status, percent)

plot_pasc90_hz <- ggplot(pasc_90_hz,
       aes(y = Phenotype, x = percent, color = status, label = count)) + 
  geom_point() +  
  facet_wrap(~group, scales = "free_y", ncol = 1) + 
  theme_bw() 

ggplotly(plot_pasc90_hz) %>% layout(
      margin = list(b = 50)) # to fully display the x and y axis labels

datatable(pasc_90_hz %>% arrange(desc(percent)), filter = "top")
```

### Identify patients with new-onset symptoms

For this analysis, we will be conservative and roll-up PheCodes when identifying a patient who has "new onset". Patients will be considered 'new onset', if their first occurrence of a rolled-up PheCode occurs after 90 days

```{r}
# roll up codes; identify patients with first occurence >= 90 days
covid_phecodes_roll_train_hz <-  train %>% 
  inner_join(., neuro_phecodes_icd %>% 
               rename(concept_code = 'ICD10'),
             by = "concept_code") %>% 
  mutate(PheCode_roll = as.numeric(substr(phecode, 1, 3))) %>% 
  group_by(patient_num, PheCode_roll) %>% 
  slice_min(order_by = days_since_admission) %>% 
  filter(days_since_admission >= 90)

# calculate counts and percents by non-rolled Phenotype/status
first_onset_hz <- covid_phecodes_roll_train_hz %>% 
  group_by(Phenotype, status) %>% 
  mutate(count = n_distinct(patient_num),
         percent = round(count/nrow(train %>% distinct(patient_num))*100,2)) %>% 
  ungroup() %>% 
  distinct(phecode, Phenotype, PheCode_roll, count, group, status, percent) 

plot_first_onset_hz <- ggplot(first_onset_hz,
       aes(y = Phenotype, x = percent, label = count, color = status)) + 
  geom_point() +  
  facet_wrap(~ group, scales = "free_y", ncol = 1) + 
  theme_bw() + 
  theme(legend.position = "none")

ggplotly(plot_first_onset_hz) %>% layout(
      margin = list(b = 50)) # to fully display the x and y axis labels

datatable(first_onset_hz %>% arrange(desc(percent)), filter = "top")
```

## Evaluate All Neuro/Mental health PheCodes

We will also include our rescued codes from before

### Frequency of ALL neuro/mental health PheCodes

```{r}
all_neuro_phecodes <- phecodes_all %>% 
  filter(group %in% c("mental disorders", "neurological")|
           phecode %in% neuro_phecodes$phecode) %>% 
  rename(Phenotype = "phecode_str",
         `ICD10.String` = "icd10cm_str",
         concept_code = "icd10cm") %>% 
  select(phecode, Phenotype, group, concept_code, ICD10.String)

all_neuro_phecodes_90_train <- train %>% 
  merge(., all_neuro_phecodes,
              by = "concept_code") %>% 
  filter(days_since_admission >= 90)
```

#### Prevalent PheCodes after 90 days

```{r fig.height=45, fig.width=10}
all_neuro_phecodes_90 <- all_neuro_phecodes_90_train %>% 
  group_by(Phenotype, status) %>% 
  mutate(count = n_distinct(patient_num),
         percent = round(count/nrow(train %>% distinct(patient_num))*100,2)) %>% 
  ungroup() %>% 
  distinct(phecode, Phenotype, count, group, status, percent)

plot_all_neuro_phecodes_90 <- ggplot(all_neuro_phecodes_90,
       aes(y = Phenotype, x = percent, color = status, label = count)) + 
  geom_point() +  
  facet_wrap(~group, scales = "free_y", ncol = 1) + 
  theme_bw() 

ggplotly(plot_all_neuro_phecodes_90) %>% layout(
      margin = list(b = 50)) # to fully display the x and y axis labels

datatable(all_neuro_phecodes_90 %>% arrange(desc(percent)), filter = "top")
```

### Identify patients with new-onset symptoms

```{r}
all_neuro_covid_phecodes_train_roll <-  train %>% 
  inner_join(., all_neuro_phecodes,
             by = "concept_code") %>% 
  mutate(PheCode_roll = as.numeric(substr(phecode, 1, 3))) %>% 
  group_by(patient_num, PheCode_roll) %>% 
  slice_min(order_by = days_since_admission) %>% 
  filter(days_since_admission >= 90) 

first_onset_all_neuro <- all_neuro_covid_phecodes_train_roll %>% 
  group_by(Phenotype, status) %>% 
  mutate(count = n_distinct(patient_num),
         percent = round(count/nrow(train %>% distinct(patient_num))*100,2)) %>% 
  ungroup() %>% 
  distinct(phecode, PheCode_roll, Phenotype, count, group, status, percent) 

plot_first_onset_all_neuro <- ggplot(first_onset_all_neuro,
       aes(y = Phenotype, x = percent, label = count, color = status)) + 
  geom_point() +  
  facet_wrap(~ group, scales = "free_y", ncol = 1) + 
  theme_bw() + 
  theme(legend.position = "none")

ggplotly(plot_first_onset_all_neuro) %>% layout(
      margin = list(b = 50)) # to fully display the x and y axis labels

datatable(first_onset_all_neuro, filter = "top")
```

# Patient PheCode vectors

We will create a few sets of patient vectors for both the 1) HZ and the 2) all neuro/mental health patient sets:

1.  Vectors indicting patients with the code at any time 90 days after hospitalization

2.  Vectors indicating patients with "new onset" at least 90 days after index hospital date

3.  Vectors indicating patients with "new onset" at least 90 days after index hospital date, where each PheCode is rolled up (this helps reduce dimensionality of the vectors)

## Create list of Rolled-up PheCodes

We will take the Phenotype name from the first level of the PheCode hierarchy

```{r}
all_neuro_phecode_rolled_list <- all_neuro_phecodes %>% 
  mutate(PheCode_roll = as.numeric(substr(phecode, 1, 3))) %>% 
  distinct(phecode, PheCode_roll, Phenotype, group) %>% 
  arrange(phecode) %>% 
  group_by(PheCode_roll) %>% 
  mutate(Phenotype_roll = first(Phenotype)) %>% 
  distinct(Phenotype, phecode, PheCode_roll, Phenotype_roll)
```

## Create patient vectors

```{r}
# for every patient we will create a vector with all codes of interest
# one-hot encode

## how many unique codes in each dataset
# HZ analysis
nrow(pasc_90_hz %>% distinct(phecode))
nrow(first_onset_hz %>% distinct(phecode))
nrow(first_onset_hz %>% distinct(PheCode_roll))
# all neuro/mental health codes + rescued codes
nrow(all_neuro_phecodes_90 %>% distinct(phecode))
nrow(first_onset_all_neuro %>% distinct(phecode))
nrow(first_onset_all_neuro %>% distinct(PheCode_roll))

# function to create vectors of interest
# `phecode_column` is the variable (Phenotype, phecode, or Phenotype_roll/PheCode_roll) that will be one-hot encoded/used as the column names

create_vector <- function(df, phecode_column) {
  
  vector <- df %>%
    distinct(patient_num, df[paste0(phecode_column)]) %>%
    mutate(present = 1) %>%
    pivot_wider(names_from = paste0(phecode_column), values_from = present) 
  
  # handle missing value
  vector[is.na(vector)] <- 0
  
  # add admission status
  vector <- vector %>% 
    left_join(., train %>% 
                distinct(patient_num, status), 
              by = "patient_num") %>% 
    select(patient_num, status, everything())
  
  return(vector)
}


vector_hz1 <- create_vector(pasc_90_train_hz, "Phenotype")
vector_hz2_first_onset <- create_vector(covid_phecodes_roll_train_hz, "Phenotype")
vector_hz3_first_onset_roll <- covid_phecodes_roll_train_hz %>% 
  left_join(., all_neuro_phecode_rolled_list, by= "phecode") %>% 
  create_vector(., "Phenotype_roll")

vector_all_neuro1 <- create_vector(all_neuro_phecodes_90_train, "Phenotype")
vector_all_neuro2_first_onset <- create_vector(all_neuro_covid_phecodes_train_roll, "Phenotype")
vector_all_neuro3_first_onset_roll <- all_neuro_covid_phecodes_train_roll %>% 
  left_join(., all_neuro_phecode_rolled_list, by= "phecode") %>% 
  create_vector(., "Phenotype_roll")


# sanity check whether any patients have no neuro codes
# if all 0s, the most could be 34
test = vector_hz1
test$sum <- rowSums(test == 0) 
max(test$sum)

# most could be 35
test2 = vector_hz2_first_onset
test2$sum <- rowSums(test2 == 0)
max(test2$sum)

# most could be 21
test3 = vector_hz3_first_onset_roll
test3$sum <- rowSums(test3 == 0)
max(test3$sum)

# most could be 160
test4 = vector_all_neuro1
test4$sum <- rowSums(test4 == 0)
max(test4$sum)

# most could be 159
test5 = vector_all_neuro2_first_onset
test5$sum <- rowSums(test5 == 0)
max(test5$sum)

# most could be 59
test6 = vector_all_neuro3_first_onset_roll
test6$sum <- rowSums(test6 == 0)
max(test6$sum)

rm(test, test2, test3, test4, test5, test6)
```

### Save patient vectors

These will be used for clustering

```{r}
pasc_vectors <- list(vector_hz1 = vector_hz1, 
                    vector_hz2_first_onset = vector_hz2_first_onset, 
                    vector_hz3_first_onset_roll = vector_hz3_first_onset_roll, 
                    vector_all_neuro1 = vector_all_neuro1,
                    vector_all_neuro2_first_onset = vector_all_neuro2_first_onset, 
                    vector_all_neuro3_first_onset_roll = vector_all_neuro3_first_onset_roll)

save(pasc_vectors, file = "results/pasc_vectors.rda")
```

### Save the tables with the frequency of each PheCode

```{r}
phecode_counts.rda <- list(pasc_90_hz = pasc_90_hz,
                         covid_phecodes_roll_train_hz = covid_phecodes_roll_train_hz, 
                         all_neuro_phecodes_90_train = all_neuro_phecodes_90_train,
                         all_neuro_covid_phecodes_train_roll = all_neuro_covid_phecodes_train_roll)


save(phecode_counts.rda, file = "results/phecode_counts.rda")
```
